A scalable semi-Lagrangian method on octree meshes

### 2012-11-19 doc/levelset/level_set_method_on_quadtree_JCP_rebuttal.tex

%-------------------------------------------------------------------------------
%
% pre-Header
%
%
%-------------------------------------------------------------------------------
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{times}
\usepackage{geometry}
\usepackage{color}
\usepackage{multirow}
\usepackage{bigstrut}

\geometry{verbose,letterpaper,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\DeclareMathOperator*{\avg}{average}
\usepackage{babel}

\begin{document}
\newtheorem{thm}{Theorem}
\newtheorem{cor}{Corollary}
\newtheorem{argm}{Argument}
\newtheorem{defn}{Definition}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{definition}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{example}[1][Example]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{remark}[1][Remark]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newcommand{\qed}{\nobreak \ifvmode \relax \else
\ifdim\lastskip<1.5em \hskip-\lastskip
\hskip1.5em plus0em minus0.5em \fi \nobreak
\vrule height0.75em width0.5em depth0.25em\fi}

\newcommand{\etal}{\textit{et al. }}
\newcommand{\TT}{\times 10^}

%-------------------------------------------------------------------------------
%
% Title
%
%-------------------------------------------------------------------------------
\title{A Second Order Accurate Level Set Method on Non-Graded Adaptive Cartesian Grids}
\author{Chohong Min \thanks{Mathematics Department, University of California, Santa Barbara, CA 93106.} \hspace{2cm}
Fr\'{e}d\'{e}ric Gibou \thanks{Mechanical Engineering Department \& Computer Science Department,
University of California, Santa Barbara, CA 93106.}} \maketitle
\begin{abstract}
We present a level set method on non-graded adaptive Cartesian grids, i.e.
grids for which the ratio between adjacent cells is not constrained. We use
quadtree and octree data structures to represent the grid and a simple
algorithm to generate a mesh with the finest resolution at the interface.
In particular, we present (1) a locally third order accurate
reinitialization scheme that transforms an arbitrary level set function
into a signed distance function, (2) a second order accurate
semi-Lagrangian methods to evolve the linear level set advection equation
under an externally generated velocity field, (3) a second order accurate
upwind method to evolve the nonlinear level set equation under a normal
velocity as well as to extrapolate scalar quantities across an interface in
the normal direction, and (4) a semi-implicit scheme to evolve the
interface under mean curvature. Combined, we obtain a level set method on
adaptive Cartesian grids with a negligible amount of mass loss. We propose
numerical examples in two and three spatial dimensions to demonstrate the
accuracy of the method.
\end{abstract}

%-------------------------------------------------------------------------------
%
% Introduction
%
%-------------------------------------------------------------------------------
\section{Introduction}

Many problems in science and engineering can be described by a moving free
boundary model. Examples include free surface flows, Stefan problems and
multiphase flows to cite a few. The difficulty in solving these problems
stems from the fact that: First, they involve dissimilar length scales.
Second, the boundary position must be computed as part of the solution
process. Third, the interface may be expected to undergo complex
topological changes, such as the merging or the pinching of two fronts.
Numerically, the interface that separates the two phases can be either
explicitly tracked or implicitly captured. Several classes of successful
methods exist with their own virtues and drawbacks.

Volume of Fluid methods
\cite{benson:1992:hydrocodes,benson:2002:VOF_Review,DeBar:1974:KRAKEN,Hirt_Nichols:1981:VOF,Noh_Woodward:1976:SLIC,Youngs:1984:VOF}
have the advantage of being volume preserving since the mass fraction in
each cell is being tracked. However, it is often difficult to extract
geometrical properties such as curvatures due to the fact that it is
challenging or even impossible to reconstruct a smooth enough function from
the mass fractions alone. We note however that some recent improvement in
interface reconstruction can be found in \cite{Dyadechko:2006:MOF}.

The main advantage of an explicit approach, e.g. front tracking
\cite{glimm:1999:fronttracking, Juric:1996:Dendritic_Solidification,
Juric:1998:Boiling_Flows, Tryggvason:2001:Front_Tracking_Flows_Review}, is
its accuracy. The main disadvantage is that additional treatments are
needed for handling changes in the interface's topology. In turn, the
explicit treatment of connectivity makes the method challenging to extend
to three spatial dimensions. While researchers have produced remarkable
results for a wide variety of applications using front tracking techniques,
these difficulties make this approach not ideally suited for studying
interface problems with changes in topology. Implicit representations such
as the level set method or the phase-field method represent the front as an
isocontour of a continuous function. Topological changes are consequently
handled in a straightforward fashion, and thus these methods are readily
implemented in both two and three spatial dimensions.

The main idea behind the phase-field method is to distinguish between
phases with an order parameter (or phase-field) that is constant within
each phase but varies smoothly across an interfacial region of finite
thickness. The dynamics of the phase-field is then coupled to that of the
solution in such a way that it tracks the interface motion and approximates
the sharp interface limit when the order parameter vanishes. Phase-field
methods are very popular techniques for simulating dendritic growth for
example and have produced accurate quantitative results, e.g.
\cite{Langer:1986:Phase_Field, Karma_Rappel:1997:Phase_Field_2D_3D,
Karma:2001:Phase_Field_Different_Diffusion,
Nestler:2005:Phase_Field_FEM_FDM, Schmidt:1996:CTD}. However, these methods
suffer from their own limitations: Phase-field methods have only an
approximate representation of the front location and thus the
discretization of the diffusion field is less accurate near the front,
resembling an enthalpy method \cite{chorin:1967:ANM}. Another consequence
is the stringent time step restriction imposed by such methods. Karma and
Rappel \cite{Karma_Rappel:1996:Phase_Field_Kinetics} have developed a
thin-interface limit of the phase-field model with a significant
improvement of the capillary length to interface thickness ratio
constraint; however, the time step restriction is still on the order of the
microscopic capillary length. Another disadvantage is the potential
difficulty in relating the phase-field parameters to the physical
parameters \cite{Wheeler:1993:Handbook}, although some progress is being
made for some wider class of problems
\cite{Elder_Provatas:2001:Phase_Field_Sharp_Interface}.

The main difference between the phase-field method and the level set approach
\cite{osher:1988:originallevelset, sethian:1999:levelset, osher:2002:levelset} is that the level
set method is a sharp interface model. The level set can therefore be used to exactly locate the
interface in order to apply discretizations that depend on the exact interface location.
Consequently, the sharp interface equation can be solved directly with no need for asymptotic
analysis, which makes the method potentially more attractive in developing general tool box
software for a wide range of applications. Another advantage is that only the standard time step
restrictions for stability and consistency are required, making the method significantly more
efficient. Level set methods have been extremely successful on uniform grids in the study of
physical problems such as compressible flows, incompressible flows, multiphase flows (see e.g.
\cite{osher:2002:levelset, sethian:1999:levelset} and the references therein), Epitaxial growth
(see e.g.\cite{Caflisch:1999:Island_Dynamics, Gibou:2001:Rate_Equations_Scaling_Laws,
Gibou:2001:Rate_Equations, Ratsch:2001:Island_Dynamics_Review} and the references therein) or in
image processing (see e.g. \cite{osher:2003:levelset} and the references therein). One of the main
problem of the level set method, namely its mass loss, has been partially solved with the advent of
the particle level set method of Enright \etal \cite{Enright:2002:PLS}. Within this method, the
interface is captured by the level set method and massless particles are added in order to reduce
the mass loss. The massless particles are also used in the reinitialization process for obtaining
smoother results for the reinitialized level set function. However, the use of particles adds to
the CPU and the memory requirement and cannot be applied for flows producing shocks. Rather
recently, there has been a thrust in developing level set methods on adaptive Cartesian grids. For
example Losasso \etal \cite{losasso:2004:octree} presented a particle level set based method to
simulate free surface flows on non-graded Cartesian grids. Within this method, the interface
between the liquid and the air is captured by the particle level set on a non graded octree data
structure. Other interesting work on adaptive level-set methods can be found in
\cite{droske01adaptive} and \cite{Milne:1995:thesis}.

In this paper, we present a general particle-less level set method on
non-graded Cartesian grids that produces a negligible amount of mass loss.
We apply this method to the level set evolution (1) with an externally
generated velocity field, (2) in the normal direction and (3) under mean
curvature. We also present a locally third order accurate reinitialization
scheme that transforms an arbitrary function into a sign distance function
as well as standard techniques to extrapolate a scalar quantities across an
interface in its normal direction.

\section{The Level Set Method}
The level set method, introduced by Osher and Sethian
\cite{osher:1988:originallevelset} describes a curve in two spatial
dimensions or a surface in three spatial dimensions by the zero-contour of
a higher dimensional function $\phi$, called the level set function. For
example, in two spatial dimension, a curve is define by $\{(x, y):
\phi(x,y)=0\}$. Under a velocity field $V$, the interface deforms according
to the level set equation
\begin{eqnarray}
\phi_t+V \cdot \nabla \phi = 0. \label{eq_levelset}
\end{eqnarray}
To keep the values of $ \phi $ close to those of a signed distance
function, i.e. $ |\nabla \phi| = 1 $, the reinitialization equation
introduced in Sussman \etal \cite{Sussman:1994:levelsetincompressible}
\begin{eqnarray}
\phi_{\tau} + S(\phi_o) \left( |\nabla \phi| - 1 \right) =0 \label{reinit}
\end{eqnarray}
is traditionally iterated for a few steps in fictitious time, $ \tau $.
Here $S(\phi_o)$ is a smoothed out sign function. The level set function is
used to compute the normal $$ \vec{n} = \nabla \phi/|\nabla \phi|, $$ and
the mean curvature $$\kappa = \nabla \cdot \vec{n}. $$ We refer the
interested readers to the book by Osher and Fedkiw
\cite{osher:2002:levelset} as well to the book by Sethian
\cite{sethian:1999:levelset} for more details on the level set method.

%-------------------------------------------------------------------------------
%
% Spatial Discretization
%
%-------------------------------------------------------------------------------
\section{Spatial Discretization and Refinement Criterion \label{sec_spatial_discretization}}

We use a standard quadtree (resp. octree) data structure to represent the
spatial discretization of the physical domain in two (resp. three) spatial
dimensions as depicted in figure \ref{fig_quadtree_description}: Initially
the root of the tree is associated with the entire domain, then we
recursively split each cell into four children until the desired level of
detail is achieved. This is done similarly in three spatial dimensions,
except that cells are split into eight cubes (children). We refer the
reader to the books of Samet \cite{Samet:1990:AOS,Samet:1989:TDA} for more
details on quadtree/octree data structures.

By definition, the difference of level between a parent cell and its direct
descendant is one. The level is then incremented by one for each new
generation of children.  A tree in which the difference of level between
adjacent cells is at most one is called a graded tree. Meshes associated
with graded trees are often used in the case of finite element methods in
order to produce procedures that are easier to implement. Graded Cartesian
grids are also used in the case of finite difference schemes, see for
example the work of Popinet \cite{Popinet:2003:Gerris} for the study of
incompressible flows. Graded meshes impose that extra grid cells must be
added in regions where they are not necessarily needed, consuming some
computational resources that cannot be spent elsewhere, eventually limiting
the highest level of detail that can be achieved. Moore
\cite{Moore:1995:The_Cost_of_Balancing_Quadtrees} demonstrates that the
cost of transforming an arbitrary quadtree into a graded quadtree could
involve 8 times as many grid nodes. Weiser \cite{Weiser:1981:thesis}
proposed a rough estimate for the three dimensional case and concluded that
as much as 71 times as many grid nodes could be needed for balancing
octrees. These estimates clearly represent the worse case scenarios that
seldom exist in practical simulations. However, there is still a non
negligible difference between graded and non graded grids. In addition, not
imposing any constraint on the difference of level between two adjacent
cells allows for easier/faster adaptive mesh generations.

\begin{figure}
\begin{center}\includegraphics[scale=0.3]{quadtree_description.eps}
\includegraphics[scale=0.3]{quadtree_description_2.eps}\end{center}
\caption{Discretization of a two dimensional domain (left) and its quadtree
representation (right). The entire domain corresponds to the root of the
tree (level 0). Then each cell can be recursively subdivided further into
four children. In this example, the tree is ungraded since the difference
of level between cells exceeds one.\label{fig_quadtree_description}}
\end{figure}

In this work we choose to impose that the finest cells lie on the
interface, since it is the region of interest for the level set method. In
order to generate adaptive Cartesian grids, one can use the signed distance
function to the interface along with the Whitney decomposition, as first
proposed by Strain in \cite{strain:1999:tree}. Simply stated, one
\textit{"splits any cell whose edge length exceeds its distance to the
interface"}. For a general function $\phi:\mathbb{R}^n\to\mathbb{R}$ with
Lipschitz constant $Lip(\phi)$, the Whitney decomposition was extended in
Min \cite{min:2004:lls}: Starting from a root cell split any cell $C$ if
$$\min_{v\in \textrm{vertices}(C)}|\phi(v)| \le \textrm{Lip}(\phi)\cdot
\textrm{diag-size}(C),$$ where diag-size($C$) refers to the length of the
diagonal of the current cell $C$ and $v$ refers to a vertex (node) of the
current cell.

%-------------------------------------------------------------------------------
%
% T-junction Nodes
%
%-------------------------------------------------------------------------------
\section{Finite Difference Discretizations\label{sec_finite_differences}}
\begin{figure}
\begin{center}\includegraphics[scale=0.45]{neighbors_of_node_quadtree.eps}
\end{center}
\caption{Neighboring nodes of a T-junction node,
$v_0$.\label{fig_T_junction_quadtree}}
\end{figure}

In the case of non regular Cartesian grids, the main difficulty comes from
deriving discretizations at T-junction nodes, i.e. nodes for which there is
a missing neighboring node in one of the Cartesian directions. For example
figure \ref{fig_T_junction_quadtree} depicts a T-junction node $v_0$, with
three neighboring nodes $v_1$, $v_2$ and $v_3$ aligned in the Cartesian
directions and one ghost neighboring node $v_4$ replacing the missing grid
node in the positive Cartesian direction. The value of a node-sampled
function $\phi:\{v_i\}\to\mathbb{R}$ at the ghost node $v_4$ could for
example be define by linear interpolation:
\begin{equation}
\phi_4^G = \frac{\phi_5 s_6+\phi_6 s_5}{s_5+s_6}. \label{eq_bilinear_interpolation_ghost_node_in_quadtree}
\end{equation}
However, instead of using this second order accurate interpolation, one can
instead use the following third order accurate interpolation: First, note
that a simple Taylor expansion demonstrates that the interpolation error in
equation \ref{eq_bilinear_interpolation_ghost_node_in_quadtree} is given
by:
\begin{equation}
\phi_4^G = \frac{\phi_5 s_6+\phi_6
s_5}{s_5+s_6}=\phi(v_4)+\frac{s_5s_6}{2}\phi_{yy}(v_0) + O(\triangle
x_{\textrm{smallest}})^3, \label{Taylor_Quadtree}
\end{equation}
where $\triangle x_{\textrm{smallest}}$ is the size of the smallest grid
cell with vertex $v_0$. The term $\phi_{yy}(v_0)$ can be approximated using
the standard first order accurate discretization
$\frac{2}{s_2+s_3}\left(\frac{\phi_2-\phi_0}{s_2}
+\frac{\phi_3-\phi_0}{s_3}\right)$ and cancelled out in equation
\ref{Taylor_Quadtree} to give:
\begin{equation}
\phi_4^G = \frac{\phi_5 s_6+\phi_6 s_5}{s_5+s_6}-\frac{s_5s_6}{s_2+s_3}\left(\frac{\phi_2-\phi_0}{s_2} +\frac{\phi_3-\phi_0}{s_3}\right). \label{eq_quadratic_interpolation_ghost_node_in_quadtree}
\end{equation}
We also point out that this interpolation only uses the node values of the
cells adjacent to $v_0$, which is particularly beneficial since access to
cells not immediately adjacent to the current cell is more difficult and
could add on CPU time and/or memory requirement.

\begin{figure}
\begin{center}\includegraphics[
  scale=0.45]{neighbors_of_node_octree.eps}
\end{center}
\caption{Neighboring vertices of a vertex three spatial
dimensions.\label{fig_T_junction_octree}}
\end{figure}

### 2012-11-19 doc/levelset/level_set_method_on_quadtree_JCP_rebuttal.tex

In three spatial dimensions, similar interpolation procedures can be used
to define the value of $\phi$ at ghost nodes. Referring to figure
\ref{fig_T_junction_octree}, a T-junction node $v_0$ has four regular
neighboring nodes and two ghost nodes. The values of a node-sampled
function $\phi:\{v_i\}\to\mathbb{R}$ at the ghost nodes $v_4$ and $v_5$ can
be defined by second order linear and bilinear interpolations as:
\begin{equation}
\begin{split}
\phi_4^G &= \frac{s_7\phi_8+s_8\phi_7}{s_7+s_8}, \\
\phi_5^G &= \frac{ s_{11}s_{12}\phi_{11}+ s_{11}s_{ 9}\phi_{12}+
s_{10}s_{12}\phi_{ 9}+ s_{10}s_{9}\phi_{10} }
{(s_{10}+s_{11})(s_{9}+s_{12})}.
\end{split}
\label{eq_linear_interpolation_ghost_node_in_octree}
\end{equation}
As in the case of quadtrees, third order accurate interpolations can be
derived by cancelling out the second order derivatives in the error term to
arrive at:
\begin{equation}
\begin{split}
\phi_4^G &= \frac{s_7\phi_8+s_8\phi_7}{s_7+s_8}
           - \frac{s_7s_8}{s_3+s_6}
            \left( \frac{\phi_3-\phi_0}{s_3} +
                   \frac{\phi_6-\phi_0}{s_6}
            \right), \\
\phi_5^G &= \frac{
s_{11}s_{12}\phi_{11}+
s_{11}s_{ 9}\phi_{12}+
s_{10}s_{12}\phi_{ 9}+
s_{10}s_{9}\phi_{10}
}{(s_{10}+s_{11})(s_{9}+s_{12})} \\
          &\quad - \frac{s_{10}s_{11}}{s_3+s_6}
            \left( \frac{\phi_3-\phi_0}{s_3} +
                   \frac{\phi_6-\phi_0}{s_6}
            \right) \\
          &\quad - \frac{s_{9}s_{12}}{s_1+s_4}
            \left( \frac{\phi_1-\phi_0}{s_1} +
                   \frac{\phi_4^G-\phi_0}{s_4}
            \right).
\end{split}
\label{eq_quadratic_interpolation_ghost_node_in_octree}
\end{equation}
We emphasize that figure \ref{fig_T_junction_octree} represents the general
configuration of neighboring nodes in the case of an octree as described in
Min \etal \cite{Min::Poisson_AMR_Second_Order}.

The third order interpolations defined above allow us to treat T-junction
nodes in a same fashion as a regular node, up to third order accuracy.
Here, we refer to a regular node as a node for which all the neighboring
nodes in the Cartesian directions exist. Therefore, we can then define
finite differences for $\phi_x$, $\phi_y$, $\phi_z$, $\phi_{xx}$,
$\phi_{yy}$ and $\phi_{zz}$ at every nodes using standard finite difference
formulas in a dimension by dimension framework. For example, refering to
figure \ref{fig_one_dimensional_adaptive_grid}, we use the standard
discretization for $\phi_x$ and $\phi_{xx}$, namely the central difference
formulas:
\begin{equation}
\begin{split}D^0_x\phi_0=\frac{\phi_2-\phi_0}{s_2}\cdot\frac{s_1}{s_1+s_2} + \frac{\phi_0-\phi_1}{s_1}\cdot\frac{s_2}{s_1+s_2},\\
D^0_{xx}\phi_0=\frac{\phi_2-\phi_0}{s_2}\cdot\frac{2}{s_1+s_2} - \frac{\phi_0-\phi_1}{s_1}\cdot\frac{2}{s_1+s_2},
\end{split}
\label{Central_Differencing}
\end{equation}
the forward and backward first order accurate approximations of the first order derivatives:
\begin{equation}
\begin{split}
D^+_x\phi_0 &= \frac{\phi_2-\phi_0}{s_2},\\
D^-_x\phi_0 &= \frac{\phi_0-\phi_1}{s_1},
\end{split}
\label{one_sided_1_fdm}
\end{equation}
and the second order accurate approximations of the first order derivatives:
\begin{equation}
\begin{split}
D^+_x\phi_0 &= \frac{\phi_2-\phi_0}{s_2} - \frac{s_2}{2}\text{minmod}\left(D^0_{xx}\phi_0,D^0_{xx}\phi_2\right),\\
D^-_x\phi_0 &= \frac{\phi_0-\phi_1}{s_1} +
\frac{s_1}{2}\text{minmod}\left(D^0_{xx}\phi_0,D^0_{xx}\phi_1\right),
\end{split}
\label{one_sided_2_fdm}
\end{equation}
where we use the minmod slope limiter \cite{Shu:1988:ENO,liu:1996:WENO}
because it produces more stable results in region where $\phi$ might
present kinks. Similarly, approximations for first and second order
derivatives are obtained in the $y$ and $z$ directions.

\begin{figure}
\begin{center}
\includegraphics[scale=0.9]{one_dimensional_adaptive_grid.eps}
\end{center}
\caption{One dimensional adaptive grid}
\label{fig_one_dimensional_adaptive_grid}
\end{figure}
%-------------------------------------------------------------------------------
%
% Interpolations
%
%-------------------------------------------------------------------------------
\section{Interpolation Procedures\label{sec_interpolation}}
Some reserve must be provided to define data anywhere in a cell, for
example in order to use semi-Lagrangian methods (see section
\ref{sec_semi_lagrangian}). As pointed out in Strain
\cite{Strain:2000:FSL}, the most natural choice of interpolation in
quadtree (resp. octree) data structures is the piecewise bilinear (resp.
trilinear) interpolation: Consider a cell $C$ with dimensions $[0, 1]^2$,
the bilinear interpolation at a point $x \in C$ using the values at the
nodes reads:
\begin{equation}
\begin{split}
\phi(x,y) &= \phi(0,0)(1-x)(1-y) \\
          &+ \phi(0,1)(1-x)(\quad\,\,\,\,\,y) \\
          &+ \phi(1,0)(\quad\,\,\,\,\,x)(1-y) \\
          &+ \phi(1,1)(\quad\,\,\,\,\,x)(\quad\,\,\,\,\,y)
\end{split}
\label{Bilinear_Interpolation}
\end{equation}
Quadratic interpolation can also easily be constructed using the data from the parent cell: Since
the parent cell of any current cell of a quadtree (resp. octree) owns $2\times 2$ children cells
(resp. $2\times 2\times 2$) and $3\times 3$ nodes (resp. $3\times 3\times 3$), one can defined the
multidimensional Lagrange quadratic interpolation on the parent cell. For example in the case of a
cell $[-1, 1]^2$ in a quadtree, we can define the Lagrange interpolation as:
\begin{equation*}
\begin{aligned}
\phi(x,y) &= \phi(\textrm{-1},\textrm{-1})\frac{x(x-1)}2\frac{y(y-1)}2 \\
          &+ \phi(\textrm{-1},\,\,      0)\frac{x(x-1)}2(y^2-1)        \\
          &+ \phi(\textrm{-1},\,\,      1)\frac{x(x-1)}2\frac{y(y+1)}2
\end{aligned}
\begin{aligned}
+ \phi(0,\textrm{-1})(x^2-1)\frac{y(y-1)}2 &+ \phi(1,\textrm{-1})\frac{x(x+1)}2\frac{y(y-1)}2 \\
+ \phi(0,\,\,      0)(x^2-1)(y^2-1)        &+ \phi(1,\,\,     0)\frac{x(x+1)}2(y^2-1)        \\
+ \phi(0,\,\,      1)(x^2-1)\frac{y(y+1)}2 &+ \phi(1,\,\,
1)\frac{x(x+1)}2\frac{y(y+1)}2.
\end{aligned}
\end{equation*}
However, this interpolation procedure is sensitive to nearby
discontinuities, e.g. near kinks. We therefore prefer to define a quadratic
interpolation by correcting equation (\ref{Bilinear_Interpolation}) using
second order derivatives. For a cell $[0, 1]^2$, we have:
\begin{equation}
\begin{aligned}
\phi(x,y) &= \phi(0,0)(1-x)(1-y) \\
          &+ \phi(0,1)(1-x)(\quad\,\,\,\,\,y) \\
          &+ \phi(1,0)(\quad\,\,\,\,\,x)(1-y) \\
          &+ \phi(1,1)(\quad\,\,\,\,\,x)(\quad\,\,\,\,\,y)
\end{aligned}
\begin{aligned}
\\
\\
\\        &- \phi_{xx}\frac{x(1-x)}2
           - \phi_{yy}\frac{y(1-y)}2,
\end{aligned}\label{Stabilized_Interpolation1}
\end{equation}
where we define
\begin{equation}
\begin{aligned}
\phi_{xx}&=\min_{v\in \textrm{vertices}(C)}(|D^0_{xx}\phi_{v}|),\\
\phi_{yy}&=\min_{v\in \textrm{vertices}(C)}(|D^0_{yy}\phi_{v}|).
\end{aligned}\label{Stabilized_Interpolation2}
\end{equation}
Since a distant function is piecewise differentiable in general, the choice of the smallest in
absolute value enhances the numerical stability of the interpolation.

%-------------------------------------------------------------------------------
%
% Reinitialization
%
%-------------------------------------------------------------------------------
\section{Reinitialization Scheme\label{sec_reinitialization}}
In principle, the level function can be chosen as any Lipschitz continuous function. However, the
so-called signed distance function is known to produce more robust numerical results, to improve
mass conservation and to reduce errors in the computations of geometrical quantities such as the
interface curvatures. Sussman \etal proposed in \cite{sussmanfatemi:1998:redistancing} to evolve
the following partial differential equation to steady state in order to reinitialize a level set
function $\phi^0:\mathbb{R}^n\to\mathbb{R}$ into the signed distance function $\phi$:
\begin{equation}
\phi_{\tau} + \mathrm{sgn}(\phi^0)\left( |\nabla\phi|-1 \right)=0,
\label{eq_reinitialization}
\end{equation}
where $\tau$ represents the fictitious time. A standard discretization for
this equation in its semi-discrete form is given by:
\begin{equation}\frac{d\phi}{d\tau} + \mathrm{sgn}(\phi^0)
\left[ H_G\left(D_x^+\phi, D_x^-\phi,
                D_y^+\phi, D_y^-\phi\right)-1\right]=0,\label{Semi_Discrete}\end{equation}
where $\mathrm{sgn}(\phi^0)$ denotes the signum of $\phi^0$ and $H_G$ is the Godunov Hamiltonian
defined as:
\begin{equation*}
H_G(a,b,c,d)=
\begin{cases}
\sqrt{\max(|a^+|^2,|b^-|^2)+\max(|c^+|^2,|d^-|^2)} & \textrm{ if }
\mathrm{sgn}(\phi^0)\le 0 \\
\sqrt{\max(|a^-|^2,|b^+|^2)+\max(|c^-|^2,|d^+|^2)} & \textrm{ if } \mathrm{sgn}(\phi^0)>0
\end{cases},
\end{equation*}
with $a^+=\max(a,0)$ and $a^-=\min(a,0)$. The one-sided derivatives,
$D_x^{\pm}\phi$ and $D_y^{\pm}\phi$ are discretized by the second order
accurate one-sided finite differences defined in section
\ref{sec_finite_differences}. Equation (\ref{Semi_Discrete}) is evolved in
time with the TVD RK-2 method given in Shu and Osher \cite{Shu:1988:ENO}:
First define $\tilde{\phi}^{n+1}$ and $\tilde{\phi}^{n+2}$ by Euler steps
\begin{eqnarray*}
\frac{\tilde{\phi}^{n+1}-\phi^n}{\triangle \tau} &+& \mathrm{sgn}(\phi^0) \left[
H_G\left(D_x^+\phi^n, D_x^-\phi^n,
                D_y^+\phi^n, D_y^-\phi^n\right)-1\right]=0,\\
\frac{\tilde{\phi}^{n+2}-\tilde{\phi}^{n+1}}{\triangle \tau} &+& \mathrm{sgn}(\phi^0) \left[
H_G\left(D_x^+\tilde{\phi}^{n+1}, D_x^-\tilde{\phi}^{n+1},
                D_y^+\tilde{\phi}^{n+1}, D_y^-\tilde{\phi}^{n+1}\right)-1\right]=0,
\end{eqnarray*} and then define $\phi^{n+1}$ by averaging:
\begin{eqnarray*}
\phi^{n+1} &=& \frac{\phi^n + \tilde{\phi}^{n+2}}2.
\end{eqnarray*}

In order to preserve area/volume, the reinitialization procedure is
required not to move the original interface defined by $\phi_0$. In their
seminal work, Russo and Smereka \cite{Russo_Smereka} solved this problem by
simply including the initial interface location (given by $\phi_0$) in the
stencils of the one-sided derivatives. Consider the case depicted by figure
\ref{fig_one_dimensional_adaptive_grid} and suppose that
$\phi^0_0\cdot\phi^0_2<0$, i.e. the interface is located between the nodes
$v_0$ and $v_2$. The interface location $v_I$ can be calculated by finding
the root of the quadratic interpolation of $\phi^0$ on the interval
$\overline{v_0v_2}$ with the origin at the center of the interval:
\begin{equation*}
\phi^0(x)=c_2x^2+c_1x+c_2,
\text{  with }
\begin{cases}
c_2 &= \frac12\text{minmod}\left[D^0_{xx}\phi^0_0,D^0_{xx}\phi^0_2\right] \\
c_1 &= (\phi^0_2-\phi^0_0)/s_2 \\
c_0 &= (\phi^0_2+\phi^0_0)/  2 - c_2s_2^2/4
\end{cases}.
\end{equation*}
The distance $s_I$ between $v_0$ and the interface location is then defined by:
\begin{equation*}
s_I = \frac{s_2}2 +
\begin{cases}
-c_0/c_1                            & \text{ if $|c_2|<  \epsilon$ }\\
(-c_1+\sqrt{c_1^2-4c_2c_0})/(2c_2)  & \text{ if $|c_2|\ge\epsilon$ and $\phi^0_0<0$}\\
(-c_1-\sqrt{c_1^2-4c_2c_0})/(2c_2)  & \text{ if $|c_2|\ge\epsilon$ and $\phi^0_0>0$}
\end{cases}.
\end{equation*}
The calculation of $D_x^+\phi^n_0$ is then modified using the interface
location and the fact that $\phi=0$ at the interface:
\begin{equation*}
D^+_x\phi^n_0 = \frac{0-\phi_0^n}{s_I} -
\frac{s_I}{2}\text{minmod}\left(D^0_{xx}\phi^n_0,D^0_{xx}\phi^n_2\right).
\end{equation*}
We note that in the original work of Russo and Smereka
\cite{Russo_Smereka}, a cubic interpolation was employed to locate the
interface, but that the above quadratic interpolation with the minmod
operator acting on the second order derivatives proved to be more stable in
the case where the level set function presents a kink nearby. We also point
out that in the original work of \cite{Russo_Smereka}, the first order
derivative $D^+_x\phi^n_0$ was discretized as:
\begin{equation*}
D^+_x\phi^n_0 = \frac{0-\phi_0^n}{s_I} -
\frac{s_I}{2}\text{minmod}\left(D^0_{xx}\phi^n_0,D^0_{xx}\phi^n_I\right),
\end{equation*}
thus included $v_I$ in the discretization of $D^0_{xx}\phi^n_I$. However,
we found that this choice leads to unstable results when the interface is
close to grid nodes. We thus slightly changed the discretization by only
using the location of the interface in the first term in order to maintain
the location of $\phi^0$, not in the discretization of second order
derivatives. Likewise, in the case where $s_I$ is close to zero (hence
$\phi^0_0$ is close to zero) we simply set $\phi^n_0=0$ to guarantee
stability. This only introduces a negligible perturbation in the location
of the zero level set.

The same process is then applied to $D^-_x\phi$ if there is a sign change between $\phi^0_0$ and
$\phi^0_1$. The time step restriction for cells cut by the interface is then:
\begin{eqnarray}
\Delta \tau = \begin{cases}
\min(s_I,s_1,s_2)                   & \textrm{in $1D$}, \\
\min(s_I,s_1,s_2,s_3,s_4)/2         & \textrm{in $2D$}, \\
\min(s_I,s_1,s_2,s_3,s_4,s_5,s_6)/3 & \textrm{in $3D$}. \\
\end{cases}
\end{eqnarray}

\subsection{Adaptive time stepping}
We note that an adaptive time step is possible since only the steady state
of \eqref{eq_reinitialization} is sought. Since the time step restriction
is adapted for each cell, the reinitialization procedure is fast: small
cells with a stringent time step restriction are located near the interface
and therefore only a few iterations are required to reach the steady state
at those cells (characteristic information flow away \textit{from} the
interface); cells far away from the interface are large and therefore do
not require a small time step restriction. For example, consider the
example depicted in figure \ref{fig_ex_reinitialization_square}, for which
the level set function is defined initially as -1 inside a square domain
(not aligned with the grid cells) and +1 outside. This initial level set
function is therefore very far from the signed distance function that we
seek to define. However, on a grid where the smallest grid has size
$dx=1/2048$, the reinitialization procedure takes only 35 iterations to
fully converge to the signed distance function in the entire domain. In
practice, the initial level set is never that far to the signed distance
function and therefore only about 5 iterations are required regardless of
the resolution of the finest level. Figure
\ref{fig_reinitialization_error_decay} illustrates the difference in the
number of iterations required between uniform time stepping and adaptive
time stepping. In the case of a uniform time step, we take $\triangle
t=\triangle x_{\text{smallest}}/2$, with $\triangle x_{\text{smallest}}$
the size of the smallest cell.

### 2012-11-19 doc/semilagr/semilagrangian.tex
Some text
Complet basic sl algorithm

\begin{algorithm}
  \caption{Basic semi-Lagrangian procedure}
  \begin{algorithmic}
    \STATE Do something.
  \end{algorithmic}
\end{algorithm}

>>>>

\begin{algorithm}
  \caption{Basic semi-Lagrangian time step}
  \begin{algorithmic}
    \FOR{$e$ in local cells}
      \FOR{$n$ in nodes on $e$ that have not yet been processed}
        \STATE Extrapolate location of $n$ backward in time as $x_n$
	\IF{$x_n$ is located on a different processor $p_n$}
	\STATE Append $x_n$ to send buffer for $p_n$
	\ENDIF
      \ENDFOR
    \ENDFOR
    \STATE Reverse communication pattern so the receivers know the senders
    \STATE Send the non-local points $x_n$ to the target processors
    \STATE Loop over local points $x_n$ and interpolate new values at $n$
    \STATE Receive points $x_n$, interpolate, add to send buffer
    \STATE Send back the interpolated values
    \STATE Receive values for non-local points and update node values
  \end{algorithmic}
\end{algorithm}
